---
title: "Comprehensive Genomic Profiling Report"
subtitle: "TruSight Oncology 500 — ESMO 2024 Compliant"
date: today
format:
  html:
    theme:
      light: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    page-layout: article
    code-fold: true
    code-summary: "Show analysis code"
    fig-align: center
    number-sections: true
    smooth-scroll: true
    mainfont: system-ui
csl: ama.csl
params:
  sample_id: "SAMPLE_001"
  config_path: "config/default.yaml"
  reports_dir: "reports"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(gt)
  library(glue)
  library(ggplot2)
  library(ggiraph)
  library(plotly)
  library(purrr)
  library(stringr)
  library(jsonlite)
  library(here)
})

source(here::here("R/plot_helpers.R"))

# Load all stage data from RDS
data_dir <- file.path(params$reports_dir, params$sample_id, "08-report", "data")
config <- readRDS(file.path(data_dir, "config.rds"))
qc <- readRDS(file.path(data_dir, "qc.rds"))
variants <- readRDS(file.path(data_dir, "variants.rds"))
cnv_data <- readRDS(file.path(data_dir, "cnv.rds"))
fusions_data <- readRDS(file.path(data_dir, "fusions.rds"))
tmb <- readRDS(file.path(data_dir, "tmb.rds"))
msi <- readRDS(file.path(data_dir, "msi.rds"))
hrd <- readRDS(file.path(data_dir, "hrd.rds"))
oncokb <- readRDS(file.path(data_dir, "oncokb.rds"))
escat <- readRDS(file.path(data_dir, "escat.rds"))
literature <- readRDS(file.path(data_dir, "literature.rds"))

civic <- tryCatch(
  readRDS(file.path(data_dir, "civic.rds")),
  error = function(e) list(variant_evidence = tibble(), gene_summaries = list(), assertions = tibble())
)
has_civic <- nrow(civic$variant_evidence) > 0 || nrow(civic$assertions) > 0

tumor_type_label <- "High-Grade Serous Ovarian Carcinoma (HGSOC)"
```

::: {.callout-note appearance="minimal"}
**Sample:** `r params$sample_id` | **Tumor:** `r tumor_type_label` | **Panel:** TSO500 (523 genes, 1.94 Mb) | **Source:** EGA EGAD50000001451
:::

# Patient & Sample Information

```{r patient-info}
tibble(
  Field = c("Sample ID", "Tumor Type", "Panel", "Specimen Type",
            "Sequencing Platform", "Reference Genome", "Institution", "Report Date"),
  Value = c(params$sample_id, tumor_type_label, "TruSight Oncology 500 (TSO500)",
            "FFPE Tumor Tissue", "Illumina NovaSeq X", "GRCh38 (hg38)",
            config$report$institution %||% "Research Institution",
            format(Sys.time(), "%Y-%m-%d"))
) |>
  gt() |>
  cols_width(Field ~ px(200)) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = Field)) |>
  tab_options(table.width = pct(100))
```

# Quality Control

```{r qc-summary}
qc_skipped <- isTRUE(qc$skipped)
if (qc_skipped) {
  knitr::asis_output(paste0(
    '::: {.callout-note}\n',
    '**QC skipped** — Input was a VCF file. BAM-level quality metrics are not available. ',
    'To get full QC, provide a BAM file as input.\n',
    ':::\n'))
}

if (!qc_skipped) {
qc_df <- tibble(
  Metric = c("Total Reads", "Mapped Reads", "Mapping Rate", "Mean Coverage",
             "On-Target Rate", "Duplicate Rate", "Tumor Purity"),
  Value = c(
    format(qc$summary$total_reads, big.mark = ","),
    format(qc$summary$mapped_reads, big.mark = ","),
    paste0(round(qc$summary$mapping_rate * 100, 1), "%"),
    paste0(qc$summary$mean_coverage, "x"),
    paste0(round(qc$summary$on_target_rate * 100, 1), "%"),
    paste0(round(qc$summary$duplicate_rate * 100, 1), "%"),
    paste0(round(qc$summary$tumor_purity * 100, 0), "%")
  ),
  Threshold = c("-", "-", ">95%", ">200x", ">80%", "<30%", ">10%"),
  Status = c("-", "-",
    ifelse(qc$summary$mapping_rate > 0.95, "PASS", "FAIL"),
    ifelse(qc$summary$mean_coverage > 200, "PASS", "FAIL"),
    ifelse(qc$summary$on_target_rate > 0.80, "PASS", "FAIL"),
    ifelse(qc$summary$duplicate_rate < 0.30, "PASS", "FAIL"),
    ifelse(qc$summary$tumor_purity > 0.10, "PASS", "FAIL"))
)

qc_df |>
  gt() |>
  tab_style(
    style = cell_fill(color = "#c8e6c9"),
    locations = cells_body(columns = Status, rows = Status == "PASS")
  ) |>
  tab_style(
    style = cell_fill(color = "#ffcdd2"),
    locations = cells_body(columns = Status, rows = Status == "FAIL")
  ) |>
  tab_header(
    title = glue("Overall QC: {if (qc$pass) 'PASS' else 'FAIL'}")
  ) |>
  tab_options(table.width = pct(100))
} # end if (!qc_skipped)
```

## Annotation Methodology

```{r annotation-methods}
tibble(
  Component = c("Variant Calling", "Variant Annotation", "Transcript Database",
                "Functional Prediction", "Population Frequency", "Clinical Databases",
                "Clinical Actionability", "Nomenclature", "CNV Detection", "Fusion Detection",
                "Community Evidence", "Oncogenicity Classification"),
  `Tool / Database` = c("GATK Mutect2 v4.6.1", "Ensembl VEP (GRCh38)",
    "Ensembl / GENCODE canonical", "SIFT + PolyPhen-2", "gnomAD v4.0",
    "ClinVar + COSMIC", "OncoKB API (live)", "HGVS (varnomen.hgvs.org)",
    "CNVkit (hybrid mode)", "Manta SV caller",
    "CiVIC (civicdb.org)", "AMP/ASCO/CAP Four-Tier System"),
  Details = c("Tumor-only mode, panel intervals", "--everything --pick --canonical --hgvs",
    "One consequence per variant via --pick", "Missense protein impact prediction",
    "Germline filtering at AF >1%", "Pathogenicity + somatic mutation catalogs",
    "Therapeutic levels → ESCAT tiers", "Protein (p.) and coding (c.) notation",
    "Panel-optimized copy number segmentation", "Structural variants for gene fusions",
    "CiVIC GraphQL API — variant evidence, gene summaries, community-curated assertions",
    "Li et al., J Mol Diagn 2017; combines OncoKB + CiVIC + VEP + ClinVar evidence")
) |>
  gt() |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = Component)) |>
  tab_options(table.width = pct(100))
```

# Somatic Variants

`r nrow(variants |> filter(consequence != "amplification"))` somatic variants detected. Annotated with Ensembl VEP (GRCh38) using canonical transcripts and [HGVS nomenclature](https://varnomen.hgvs.org/).

```{r somatic-variants}
variants |>
  filter(consequence != "amplification") |>
  mutate(
    Gene_display = ifelse(!is.na(transcript_id),
      glue("{gene}\n{transcript_id}"), gene),
    VAF_display = ifelse(!is.na(vaf),
      glue("{round(vaf*100,1)}%\n({alt_depth}/{total_depth})"), "N/A"),
    Functional = ifelse(!is.na(sift_prediction),
      glue("{sift_prediction}\n{polyphen_prediction}"), "-")
  ) |>
  select(
    Gene = Gene_display, `HGVS Protein` = hgvsp, `HGVS Coding` = hgvsc,
    VAF = VAF_display, Consequence = consequence, Impact = impact,
    `SIFT / PolyPhen` = Functional, ClinVar = clinvar_significance,
    Classification = pathogenicity
  ) |>
  gt() |>
  tab_style(
    style = cell_fill(color = "#ffebee"),
    locations = cells_body(rows = Classification %in% c("pathogenic", "likely_pathogenic"))
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Gene)
  ) |>
  fmt_missing(everything(), missing_text = "-") |>
  tab_options(table.width = pct(100), table.font.size = "small")
```

## Mutation Confidence Ranking

Variants ranked by call confidence based on VAF relative to detection threshold (`r round(config$variant_calling$min_vaf * 100, 0)`%), sequencing depth, and database concordance.

```{r confidence-ranking}
if ("confidence" %in% names(variants)) {
  variants |>
    filter(consequence != "amplification", !is.na(confidence)) |>
    select(Gene = gene, Alteration = hgvsp, Confidence = confidence,
           Rationale = confidence_reason) |>
    gt() |>
    tab_style(
      style = cell_fill(color = "#c8e6c9"),
      locations = cells_body(columns = Confidence, rows = Confidence == "High confidence")
    ) |>
    tab_style(
      style = cell_fill(color = "#fff9c4"),
      locations = cells_body(columns = Confidence, rows = Confidence == "Uncertain")
    ) |>
    tab_style(
      style = cell_fill(color = "#ffcdd2"),
      locations = cells_body(columns = Confidence, rows = str_detect(Confidence, "Questionable"))
    ) |>
    tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = Gene)) |>
    tab_footnote("Questionable variants should be confirmed by orthogonal methods (Sanger, ddPCR).") |>
    tab_options(table.width = pct(100), table.font.size = "small")
}
```

## VAF Distribution

```{r vaf-distribution, fig.width=8, fig.height=5}
vaf_df <- variants |>
  filter(consequence != "amplification", !is.na(vaf)) |>
  select(gene, hgvsp, vaf, consequence, pathogenicity)

plot_vaf_distribution(vaf_df, config$variant_calling$min_vaf)
```

# Copy Number Alterations

```{r cnv-table}
cnv_data |>
  select(Gene = gene, Type = type, `Log2 Ratio` = log2_ratio,
         `Copy Number` = copy_number, Chromosome = chromosome,
         Start = start, End = end) |>
  gt() |>
  tab_style(
    style = cell_text(color = "#c62828", weight = "bold"),
    locations = cells_body(columns = Type, rows = Type == "AMPLIFICATION")
  ) |>
  tab_style(
    style = cell_text(color = "#1565c0", weight = "bold"),
    locations = cells_body(columns = Type, rows = Type == "DELETION")
  ) |>
  fmt_number(columns = `Log2 Ratio`, decimals = 2) |>
  fmt_integer(columns = c(Start, End)) |>
  tab_options(table.width = pct(100))
```

## Genome-Wide CNV Profile

```{r cnv-genome-interactive, fig.width=10, fig.height=5}
plot_cnv_genome_interactive(cnv_data)
```

## Genomic Overview (Circos)

```{r circos-plot, eval=nrow(cnv_data) > 0 || nrow(fusions_data) > 0, fig.width=8, fig.height=8}
circos_file <- file.path(tempdir(), paste0(params$sample_id, "_circos.png"))
plot_circos(cnv_data, fusions_data, output_file = circos_file)
knitr::include_graphics(circos_file)
```

# Gene Fusions

```{r fusions-table}
fusions_data |>
  mutate(
    Fusion = glue("{gene_a}::{gene_b}"),
    Exons = glue("exon {exon_a}::exon {exon_b}"),
    Breakpoints = glue("{chr_a}:{format(pos_a, big.mark=',')} — {chr_b}:{format(pos_b, big.mark=',')}")
  ) |>
  select(Fusion, Exons, Breakpoints, `Supporting Reads` = supporting_reads,
         Type = fusion_type, `Known Fusion` = known_fusion) |>
  gt() |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = Fusion)) |>
  tab_options(table.width = pct(100))
```

## Fusion Visualization

```{r fusion-arcs, fig.width=8, fig.height=4}
fusion_plot_df <- fusions_data |>
  select(gene_a, gene_b, supporting_reads, fusion_type, known_fusion)
plot_fusion_arcs(fusion_plot_df)
```

# Genomic Biomarkers

```{r biomarker-gauges, fig.width=10, fig.height=4}
plot_biomarker_gauges(tmb, msi, hrd)
```

## Details {.panel-tabset}

### TMB

::: {.callout-tip appearance="simple"}
**`r tmb$tmb_score` mutations/Mb — `r tmb$tmb_class`**
:::

- Variant count: `r tmb$variant_count` non-synonymous coding variants
- Panel coding size: `r tmb$panel_size_mb` Mb (TSO500)
- Classification: TMB-High ≥10, TMB-Intermediate 6–10, TMB-Low <6

### MSI

::: {.callout-tip appearance="simple"}
**`r msi$msi_score`% — `r msi$msi_status`**
:::

- Unstable sites: `r msi$unstable_sites` / `r msi$total_sites`
- Threshold: ≥`r config$biomarkers$msi$msi_h_threshold`% for MSI-H

### HRD

::: {.callout-tip appearance="simple"}
**Score `r hrd$hrd_score` — `r hrd$hrd_status`**
:::

- LOH: `r hrd$loh_score` | TAI: `r hrd$tai_score` | LST: `r hrd$lst_score`
- Threshold: ≥`r config$biomarkers$hrd$score_threshold` for HRD-positive
- *Note: Panel-based estimation — confirm with dedicated HRD assay for clinical decisions.*

# Clinical Actionability (ESCAT)

```{r escat-summary}
actionable_n <- nrow(escat |> filter(escat_tier %in% c("I", "II", "III")))
```

::: {.callout-important}
**`r actionable_n` actionable alteration(s) identified** (ESCAT Tier I–III) across `r nrow(escat)` total classified alterations.
:::

```{r escat-table}
escat_colors <- c("I" = "#d32f2f", "II" = "#f57c00", "III" = "#fbc02d",
                  "IV" = "#7cb342", "X" = "#9e9e9e")

# Determine evidence sources for each alteration
escat_display <- escat |>
  mutate(
    civic_match = if (has_civic && nrow(civic$variant_evidence) > 0) {
      gene %in% civic$variant_evidence$query_gene
    } else {
      FALSE
    },
    Sources = case_when(
      civic_match ~ "OncoKB + CiVIC",
      TRUE ~ "OncoKB"
    )
  )

escat_display |>
  select(`ESCAT Tier` = escat_tier, Gene = gene, Alteration = alteration,
         Type = type, Oncogenic = oncogenic, Level = sensitive_level,
         Sources, Description = escat_description) |>
  gt() |>
  tab_style(
    style = list(cell_fill(color = "#ffebee"), cell_text(weight = "bold")),
    locations = cells_body(rows = `ESCAT Tier` == "I")
  ) |>
  tab_style(
    style = list(cell_fill(color = "#fff3e0"), cell_text(weight = "bold")),
    locations = cells_body(rows = `ESCAT Tier` == "II")
  ) |>
  tab_style(
    style = cell_fill(color = "#fffde7"),
    locations = cells_body(rows = `ESCAT Tier` == "III")
  ) |>
  fmt_missing(everything(), missing_text = "-") |>
  tab_options(table.width = pct(100), table.font.size = "small")
```

## Community Evidence (CiVIC)

```{r civic-evidence}
if (has_civic && nrow(civic$variant_evidence) > 0) {
  civic$variant_evidence |>
    select(Gene = query_gene, Variant = query_variant,
           Type = evidence_type, Level = evidence_level,
           Direction = evidence_direction, Significance = significance,
           Disease = disease, Therapies = therapies,
           Source = source_citation) |>
    gt() |>
    tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = Gene)) |>
    sub_missing(everything(), missing_text = "-") |>
    tab_header(title = "CiVIC Variant Evidence Items") |>
    tab_footnote("Evidence from CiVIC (civicdb.org) — community-curated variant interpretations.") |>
    tab_options(table.width = pct(100), table.font.size = "small")
} else {
  knitr::asis_output('::: {.callout-note appearance="simple"}\nNo CiVIC community evidence found for the identified variants.\n:::')
}
```

### CiVIC AMP/ASCO/CAP Assertions

```{r civic-assertions}
if (has_civic && nrow(civic$assertions) > 0) {
  civic$assertions |>
    select(Gene = gene, Variant = variant_name, `AMP Level` = amp_level,
           Type = assertion_type, Direction = assertion_direction,
           Significance = significance, Disease = disease,
           Therapies = therapies, `NCCN Guideline` = nccn_guideline,
           `FDA Companion Test` = fda_companion_test) |>
    gt() |>
    tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = Gene)) |>
    sub_missing(everything(), missing_text = "-") |>
    tab_header(title = "CiVIC AMP/ASCO/CAP Assertions") |>
    tab_footnote("AMP/ASCO/CAP tier assertions curated by the CiVIC community.") |>
    tab_options(table.width = pct(100), table.font.size = "small")
} else {
  knitr::asis_output('::: {.callout-note appearance="simple"}\nNo CiVIC AMP/ASCO/CAP assertions found for the identified genes.\n:::')
}
```

# Oncogenicity Classification (AMP/ASCO/CAP)

```{r amp-table}
amp_data <- tryCatch(
  readRDS(file.path(data_dir, "amp.rds")),
  error = function(e) NULL
)

if (!is.null(amp_data) && nrow(amp_data) > 0) {
  amp_data |>
    select(Gene = gene, Alteration = hgvsp, `AMP Tier` = amp_tier,
           `AMP Level` = amp_level, Evidence = amp_evidence) |>
    gt() |>
    tab_header(title = "AMP/ASCO/CAP Somatic Variant Classification") |>
    tab_style(
      style = list(cell_fill(color = "#ffebee"), cell_text(weight = "bold")),
      locations = cells_body(rows = `AMP Tier` == "Tier I")
    ) |>
    tab_style(
      style = cell_fill(color = "#fff3e0"),
      locations = cells_body(rows = `AMP Tier` == "Tier II")
    ) |>
    tab_style(
      style = cell_fill(color = "#fffde7"),
      locations = cells_body(rows = `AMP Tier` == "Tier III")
    ) |>
    sub_missing(everything(), missing_text = "-") |>
    tab_footnote("Classification per Li et al., J Mol Diagn 2017; 19(1):4-23") |>
    tab_options(table.width = pct(100), table.font.size = "small")
} else {
  knitr::asis_output('::: {.callout-note appearance="simple"}\nAMP/ASCO/CAP classification not available.\n:::')
}
```

# Treatment-Focused Literature Evidence

Literature search identified **`r if (is.data.frame(literature$pubmed)) nrow(literature$pubmed) else 0`** PubMed and **`r if (is.data.frame(literature$scopus)) nrow(literature$scopus) else 0`** Scopus articles. Each narrative below focuses on **therapeutic implications and recommended treatment strategies** for the identified alterations.

```{r literature-narratives, results='asis'}
if (length(literature$narratives) > 0) {
  for (key in names(literature$narratives)) {
    gene_name <- str_extract(key, "^[^_]+")
    text <- literature$narratives[[key]]
    cat(glue("\n\n## {gene_name}\n\n{text}\n\n"))
  }
}
```

```{r literature-references, results='asis'}
# Build AMA-style reference list from PubMed hits
if (is.data.frame(literature$pubmed) && nrow(literature$pubmed) > 0) {
  cat("\n\n## References\n\n")
  refs <- literature$pubmed |>
    distinct(pmid, .keep_all = TRUE) |>
    head(25)

  for (i in seq_len(nrow(refs))) {
    r <- refs[i, ]
    # AMA format: Author(s). Title. Journal. Year;Volume:Pages. doi:
    first_author <- str_extract(r$authors, "^[^,]+")
    if (is.na(first_author)) first_author <- "Unknown"
    author_count <- str_count(r$authors, ",") + 1
    author_str <- if (author_count > 3) paste0(first_author, ", et al") else r$authors

    cat(glue("{i}. {author_str}. {r$title}. *{r$journal}*. {r$year}. ",
             "PMID: [{r$pmid}](https://pubmed.ncbi.nlm.nih.gov/{r$pmid}/)\n\n"))
  }
}
```

# Coverage Gaps

## Per-Gene Coverage

```{r gene-coverage-plot, eval=!isTRUE(qc$skipped), fig.width=8, fig.height=8}
if (!is.null(qc$per_gene_coverage) && nrow(qc$per_gene_coverage) > 0) {
  plot_gene_coverage(qc$per_gene_coverage, 200)
}
```

```{r coverage-gaps}
if (qc_skipped) {
  knitr::asis_output('::: {.callout-note appearance="simple"}\nCoverage gap analysis not available — input was VCF (no BAM-level data).\n:::')
} else {
gap_genes <- qc$per_gene_coverage |> filter(mean_coverage < 200)

if (nrow(gap_genes) > 0) {
  knitr::asis_output(glue(
    '::: {{.callout-warning}}\n',
    '**{nrow(gap_genes)} gene(s)** did not meet the minimum 200x coverage threshold. ',
    'Results for these genes should be interpreted with caution.\n',
    ':::\n'))

  gap_genes |>
    arrange(mean_coverage) |>
    select(Gene = gene, `Mean Coverage` = mean_coverage,
           `Min Coverage` = min_coverage, `% ≥200x` = pct_above_200x) |>
    gt() |>
    tab_style(
      style = cell_fill(color = "#ffcdd2"),
      locations = cells_body(columns = `Mean Coverage`)
    ) |>
    tab_options(table.width = pct(100))
} else {
  knitr::asis_output('::: {.callout-tip appearance="simple"}\nAll genes met the minimum 200x coverage threshold.\n:::')
}
} # end if (!qc_skipped) for coverage
```

---

::: {.callout-note appearance="minimal"}
**Disclaimer:** This report is generated for research purposes following ESMO 2024 guidelines. Clinical decisions should be confirmed by a certified molecular pathology laboratory. OncoKB annotations provided under academic license. ESCAT classifications per Mateo et al., *Ann Oncol*, 2018. CiVIC annotations sourced from the open-access community knowledgebase (civicdb.org). AMP/ASCO/CAP classification per Li et al., J Mol Diagn 2017; 19(1):4-23.

**Pipeline:** ngs-tertiary-analysis v0.2.0 | **Reference:** GRCh38 | **Panel:** TSO500 (523 genes, 1.94 Mb coding)
:::
